A full approximation scheme multilevel method \\ for solving nonlinear variational inequalities

### 2024-07-23 talk-ismp/slides.tex
many changes to make it better

@@

\title[FAS multigrid for VIs]{A full approximation scheme multilevel method \\ for solving nonlinear variational inequalities}

@@

\titlegraphic{\vspace{-18mm} \includegraphics[width=0.22\textwidth]{../talk-oxford/images/uafbw.png} \hfill \begin{minipage}{0.12\textwidth}
\includegraphics[width=\textwidth]{../talk-oxford/images/oxford.png}

@@

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

@@

\begin{frame}{example, cont.}

@@

\vspace{-1mm}
\begin{itemize}
\item the solution defines subsets of $\Omega$:
   \begin{itemize}
   \item[$\circ$] \emph{active set} $A_u = \{u = \psi\}$
   \item[$\circ$] \emph{inactive set} $R_u = \{u> \psi\}$
   \item[$\circ$] \emph{free boundary} $\Gamma_u=\partial R_u \cap \Omega$
   \end{itemize}
\item the \emph{complementarity problem} (CP) is a PDE-like strong form:
\begin{equation*}
u - \psi \ge 0, \qquad -\grad^2 u - f \ge 0, \qquad (u - \psi)(-\grad^2 u - f) = 0
\end{equation*}
\item the weak form is the VI:
    $$\int_\Omega \grad u\cdot \grad (v-u) - f (v-u) \ge 0 \quad \forall v \in \mathcal{K}$$
\end{itemize}
\end{frame}

\begin{frame}{geometrical meaning of VI}

\begin{itemize}
\item the VI gives an ``angle condition'' for the test function $v-u$
\item in optimization problems the VI is a consequence:
$$J(u) \le J(v), \, \forall v \in \mathcal{K} \qquad \implies \qquad \ip{J'(u)}{v-u} \ge 0, \, \forall v \in \mathcal{K}$$
\item for general VIs the angle condition \emph{is} the problem:
   $$\ip{F(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}$$    
\end{itemize}

\bigskip
\begin{center}
\includegraphics[width=0.4\textwidth]{figs/convexuv.png}
\end{center}
\end{frame}

\begin{frame}{general variational inequalities}

\begin{itemize}
\item let $\mathcal{K}$ be a closed and convex subset of a Banach space $\mathcal{V}$
\item suppose $F:\mathcal{K} \to \mathcal{V}'$ is a continuous operator
    \begin{itemize}
    \item[$\circ$] $F$ is generally nonlinear
    \item[$\circ$] $F$ may be defined \emph{only} on $\mathcal{K}$
    \item[$\circ$] $F$ may \emph{not}\, be the derivative of an objective function $J$
    \end{itemize}
\item the general variational inequality {\color{FireBrick} VI($F$,$\mathcal{K}$)} is
	$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \text{ for all } v \in \mathcal{K}}$$
\item when $\mathcal{K}$ is nontrivial, VI($F$,$\mathcal{K}$) is nonlinear even when $F$ is a linear operator
\end{itemize}
\end{frame}

@@

\vspace{1.5mm}
\item first-semester calculus:
    $$u = \mathop{\textnormal{argmin}}_{x\in[a,b]} f(x) \quad \iff \quad f'(u)(v-u) \ge 0 \quad \forall v \in[a,b]$$
\end{itemize}

@@

\begin{align*}
&\text{pre-smooth over fine:} & & \text{smoother on } w^h \\
&\text{restrict:}                   & &\ell^H = F^H(\iR w^h) + R\, r^h(w^h) \\
&\text{solve coarse:}                      & &F^H(w^H) = \ell^H \\
&\text{correct:}                    & &w^h \leftarrow w^h + P(w^H - \iR w^h) \\
&\text{post-smooth over fine:} & & \text{smoother on } w^h
\end{align*}

\bigskip
{\small
\begin{itemize}
\item $R: (\mathcal{V}^h)' \to (\mathcal{V}^H)'$ is \emph{canonical restriction}
\item $P: \mathcal{V}^H \to \mathcal{V}^h$ is \emph{canonical prolongation}
\item $\iR: \mathcal{V}^h \to \mathcal{V}^H$ is \emph{injection}
\item \textbf{restrict}$+$(\textbf{solve coarse})$+$\textbf{correct} \, $=$ \, \emph{FAS coarse grid correction}
\end{itemize}
}
\end{frame}

@@

\subsection{results, and an application}

\begin{frame}{classical obstacle problem by FASCD}

@@

\begin{frame}{advection-diffusion of a concentration}

\begin{itemize}
\item $u(x)$ is a concentration in $\Omega \subset \RR^d$: \qquad $\boxed{0\le u\le 1}$
\item substance moves by diffusion, advection (wind $\bm{X}(x)$), and source ($\sigma(x)$):
    $$-\eps \grad^2 u + \bm{X}\cdot \grad u = \sigma$$
\item two active sets:
    $$\underline{A}_u = \{u(x) = 0\} \hspace{22mm} \overline{A}_u = \{u(x) = 1\} \hspace{8mm}$$
\end{itemize}

@@

\begin{frame}{advection-diffusion of a concentration}

@@

\vspace{-3mm}
\hfill \tiny \emph{Bueler \& Farrell (2024)}

\medskip
\scriptsize
\begin{itemize}
\item<2> compare linear programming (Klee-Minty cube?) \dots with strong spatial correlations
\end{itemize}
\end{frame}

\begin{frame}{problem: geometry of flowing glacier ice in a climate}

\begin{itemize}
\item ``where are the glaciers?'' is a free-boundary problem
\end{itemize}

\includegraphics[width=1.02\textwidth]{../talk-oxford/images/alps-seguinot2018.png}

\vspace{-2mm}
\hfill {\tiny Seguinot et al.~(2018)}
\end{frame}

\begin{frame}{free-boundary problem: flowing glacier ice in a climate}

@@

\begin{itemize}
\item admissible surface elevations:
    $$\mathcal{K} = \left\{r \in \mathcal{V} \,:\, r \ge b\right\}$$
\item steady-state VI problem for surface elevation $s\in\mathcal{K}$:
	$$\ip{\Phi(s) - a}{r-s} \ge 0 \quad \text{ for all } r \in \mathcal{K}$$
where $\Phi(s)=- \bu|_s \cdot \bn_s$
\item in the isothermal, nonsliding, \emph{shallow ice approximation}\footnote{$=$ \emph{a highly-simplified view of conservation of momentum, in which} $(s-b)^{8/3} \in W^{1,4}(\Omega)$ \emph{so} $\mathcal{V} \stackrel{?}{=} (W^{1,4})^{3/8}$ \hfill \emph{(Jouvet \& Bueler, 2012)}} case:
\begin{align*}
\Phi(s) &= - \bu|_s \cdot \bn_s \\
        &= - \frac{\gamma}{4} (s-b)^{4} |\grad s|^{4} - \grad \cdot\left(\frac{\gamma}{5} (s-b)^{5} |\grad s|^{2} \grad s\right)
\end{align*}
\end{itemize}
\end{frame}

\begin{frame}{test case: ice sheet ($=$ big glacier) the size of Greenland}

@@

\begin{itemize}
\item optimality of FMG solver (not shown)
\item good parallel \emph{weak scaling}
    \begin{itemize}
    \item[$\circ$] each processor owns $641\times 641$ (sub) mesh
    \item[$\circ$] $P=1024$ run had $20481^2=4.1\times 10^8$ unknowns
    \item[] \dots and 88 meter resolution
    \end{itemize}
\end{itemize}

\bigskip
\centering
\includegraphics[width=0.55\textwidth]{../talk-dms/figs/siaweaktime.png}

@@

\subsection{algorithm details}

@@

\begin{itemize}
\item new algorithm (Bueler \& Farrell 2024):

{\color{FireBrick} FASCD = full approximation scheme constraint decomposition}

\bigskip
\item what is ``constraint decomposition'' in FAS\underline{CD}?
\end{itemize}
\end{frame}

\newcommand{\cK}{\mathcal{K}}

@@

\begin{itemize}
\item start with subspace decomposition over nested meshes:
    $$\Omega^j \subset \Omega^{j+1}$$
\item the FE vector spaces $\mathcal{V}^j$ over $\Omega^j$ are also nested:
    $$\mathcal{V}^j \subset \mathcal{V}^{j+1}$$
\item $\mathcal{V}^J = \sum_{i=0}^J \mathcal{V}^i$ is called a \emph{subspace decomposition} (Xu 1992)

    \begin{itemize}
    \item[$\circ$] \emph{non}-unique vector space sum
    \item[$\circ$] Xu's paper explains how to analyze linear multigrid for PDEs via subspace decomposition
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Tai's (2003) constraint decomposition \emph{non-trivially} extends subspace decomposition to convex subsets
\item suppose $\mathcal{K}^J \subset \mathcal{V}^J$ is a closed and convex subset
\begin{block}{definition}
$\ds \mathcal{K}^J = \sum_{i=0}^J \mathcal{K}^i$ \quad is a \emph{constraint decomposition} (CD) if there are closed and convex subsets $\mathcal{K}^i\subset \mathcal{V}^i$, and (nonlinear) projections $\Pi_i : \mathcal{K}^J \to \mathcal{K}^i$, so that $\ds v = \sum_{i=0}^J \Pi_i v$ and a stability condition applies (not shown)
\end{block}
\end{itemize}
\end{frame}

@@

\begin{frame}{iterations over constraint decomposition}

\begin{itemize}
\item Tai proposed abstract iterations for solving $VI(F,\ell,\mathcal{K})$ over a CD $\mathcal{K}^J = \sum_{i=0}^J \mathcal{K}^i$, specifically:

@@

\item Tai's iterations are not practical because you must evaluate all residuals on the finest level
\item two techniques make a practical solver:

    \begin{itemize}
    \item[$\circ$] \emph{defect obstacles} on each level  \hfill {\scriptsize (Gr\"aser \& Kornhuber (2009)}
    \item[$\circ$] \emph{FAS coarse corrections}  \hfill {\scriptsize (Bueler \& Farrell (2024)}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item all sets $\mathcal{U}^j$ and $\mathcal{D}^j$ are cones which include the origin
\end{itemize}
\end{frame}

@@

see (Bueler \& Farrell 2024) for:
\begin{itemize}
\item generalization to upper and lower obstacles:
    $$\mathcal{K}^J = \{\underline{\psi}^J \le v^J \le \overline{\psi}^J\}$$
\item stopping criteria
    \begin{itemize}
    \item[$\circ$] evaluates CP/KKT conditions
    \end{itemize}
\item FMG cycle
\item details of $O(m_J)$ smoother
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item FASCD = new multilevel solver for VI (free-boundary) problems
    \begin{itemize}
    \item[$\circ$] implemented in Python Firedrake (over PETSc)
    \end{itemize}
\item observed optimality in many cases
    \begin{itemize}
    \item[$\circ$] actually fast on tested nonlinear PDE problems
    \end{itemize}

\bigskip\bigskip
\item many things \textbf{{\color{FireBrick} to do}}:
    \begin{itemize}
    \item[$\circ$] implement in C inside PETSc
    \item[$\circ$] prove convergence; compare (Reusken, 1988)
    \item[$\circ$] apply to space-time (\emph{parabolic}) VI problems
    \item[$\circ$] add mesh adaptivity
    \item[$\circ$] identify smoothers for problems like elastic contact
    \item[$\circ$] include membrane stresses in glacier case
    \end{itemize}
\end{itemize}
\end{frame}

>>>>

\newcommand{\cK}{\mathcal{K}}
\newcommand{\cV}{\mathcal{V}}

@@

\newcommand{\sdoi}[1]{\,{\tiny \href{https://doi.org/#1}{doi:#1}}}

\title[FASCD = multigrid for VIs]{A full approximation scheme multilevel method \\ for solving nonlinear variational inequalities}

@@

\titlegraphic{\vspace{-18mm} \includegraphics[width=0.2\textwidth]{../talk-oxford/images/uafbw.png} \hfill \begin{minipage}{0.12\textwidth}
\includegraphics[width=\textwidth]{../talk-oxford/images/oxford.png}

@@

\begin{frame}{Outline}
  \tableofcontents

\only<1>{\phantom{\emph{warning:} no convergence theorem}}
\only<2>{\emph{warning:} no convergence theorem}
\end{frame}

@@

\begin{frame}{example \dots and VI language}

@@

\vspace{-1mm}
\begin{itemize}
\item the solution defines subsets of $\Omega$:
   \begin{itemize}
   \item[$\circ$] \emph{active set} \quad $A_u = \{u = \psi\}$
   \item[$\circ$] \emph{inactive set} \quad $R_u = \{u> \psi\}$
   \item[$\circ$] \emph{free boundary} \quad $\Gamma_u=\partial R_u \cap \Omega$
   \end{itemize}
\item the \emph{complementarity problem} (CP) is a PDE-like strong form:
\begin{equation*}
u - \psi \ge 0, \qquad -\grad^2 u - f \ge 0, \qquad (u - \psi)(-\grad^2 u - f) = 0
\end{equation*}
\item the weak form is the VI:
    $$\int_\Omega \grad u\cdot \grad (v-u) - f (v-u) \ge 0 \quad \forall v \in \mathcal{K}$$
\end{itemize}
\end{frame}

\begin{frame}{variational inequalities: general setting}

\begin{itemize}
\item $\mathcal{K}$ is a closed and convex subset of a Banach space $\mathcal{V}$
\item $F:\mathcal{K} \to \mathcal{V}'$ is a continuous operator
    \begin{itemize}
    \item[$\circ$] $F$ is generally nonlinear
    \item[$\circ$] $F$ may be defined \emph{only} on $\mathcal{K}$
    \item[$\circ$] $F$ may \emph{not}\, be the derivative of an objective function $J$
    \end{itemize}
\item the variational inequality {\color{FireBrick} VI($F$,$\mathcal{K}$)} is to find $u\in\cK$ so that
	$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \text{ for all } v \in \mathcal{K}}$$

    \begin{itemize}
    \item[$\circ$] for nontrivial $\mathcal{K}$, VI($F$,$\mathcal{K}$) is nonlinear even if $F$ is linear
    \end{itemize}
\end{itemize}
\end{frame}

@@

\vspace{1.5mm}
\item first-semester calculus:
    $$u = \mathop{\textnormal{argmin}}_{x\in[a,b]} f(x) \quad \implies \quad f'(u)(v-u) \ge 0 \quad \forall v \in[a,b]$$
\end{itemize}

@@

\begin{align*}
&\text{pre-smooth over fine:} & & \text{smoother on } w^h \\
&\text{restrict:}                   & &\ell^H = F^H(\iR w^h) + R\, r^h(w^h) \\
&\text{coarse solve:}               & &F^H(w^H) = \ell^H \\
&\text{correct:}                    & &w^h \leftarrow w^h + P(w^H - \iR w^h) \\
&\text{post-smooth over fine:} & & \text{smoother on } w^h
\end{align*}

\bigskip
{\small
\begin{itemize}
\item $R: (\mathcal{V}^h)' \to (\mathcal{V}^H)'$ is \emph{canonical restriction}
\item $P: \mathcal{V}^H \to \mathcal{V}^h$ is \emph{canonical prolongation}
\item $\iR: \mathcal{V}^h \to \mathcal{V}^H$ is \emph{injection}
\item (restrict$+$coarse solve$+$correct) \, $=$ \, \emph{FAS coarse grid correction}
\end{itemize}
}
\end{frame}

@@

\subsection{results on test problems}

\begin{frame}{test problem 1: classical obstacle problem}

@@

\begin{frame}{test problem 2: advection-diffusion}

\begin{itemize}
\item $u(x)$ is a concentration in $\Omega \subset \RR^d$: \qquad $\boxed{0\le u\le 1}$
\item concentration evolves by diffusion, advection, and source:
    $$-\eps \grad^2 u + \bm{X}\cdot \grad u = \sigma$$
\item two active sets:
    $$\underline{A}_u = \{u(x) = 0\} \hspace{22mm} \overline{A}_u = \{u(x) = 1\} \hspace{8mm}$$
\end{itemize}

@@

\begin{frame}{test problem 2: advection-diffusion}

@@

\vspace{-3mm}
\hfill \tiny \emph{Bueler \& Farrell (2024)}
\end{frame}

\subsection{an application}

\begin{frame}{``where are the glaciers?'' is a free-boundary problem}

\bigskip
\includegraphics[width=1.01\textwidth]{../talk-oxford/images/alps-seguinot2018.png}

\vspace{-2mm}
\hfill {\tiny \emph{Seguinot et al.~(2018)}}
\end{frame}

\begin{frame}{inequality-constrained problem: flowing glacier ice in a climate}

@@

\begin{itemize}
\item admissible surface elevations:
    $$\mathcal{K} = \left\{r \in \mathcal{V} \,:\, r \ge b\right\}$$
\item steady-state VI problem for surface elevation $s\in\mathcal{K}$:
	$$\ip{\Phi(s) - a}{r-s} \ge 0 \quad \text{ for all } r \in \mathcal{K}$$
where $\Phi(s)=- \bu|_s \cdot \bn_s$
\item in the isothermal, nonsliding, \emph{shallow ice approximation}\footnote{$=$ \emph{a highly-simplified view of conservation of momentum, in which} $(s-b)^{8/3} \in W^{1,4}(\Omega)$ \emph{so} $\mathcal{V} \stackrel{?}{=} (W^{1,4})^{3/8}$ \hfill \emph{Jouvet \& Bueler (2012)}} case:
\begin{align*}
\Phi(s) &= - \bu|_s \cdot \bn_s \\
        &= - \frac{\gamma}{4} (s-b)^{4} |\grad s|^{4} - \grad \cdot\left(\frac{\gamma}{5} (s-b)^{5} |\grad s|^{2} \grad s\right)
\end{align*}
\end{itemize}
\end{frame}

\begin{frame}{test case: ice sheet the size of Greenland}

@@

\begin{itemize}
\item optimality of FMG solver (left)
\item good parallel \emph{weak scaling} (right)
    \begin{itemize}
    \item[$\circ$] each processor owns $641\times 641$ (sub) mesh
    \item[$\circ$] $P=1024$ run had $20481^2=4.1\times 10^8$ unknowns
    \item[] \dots and 88 meter resolution
    \end{itemize}
\end{itemize}

\bigskip
\mbox{\includegraphics[width=0.5\textwidth]{figs/sia.png} \includegraphics[width=0.5\textwidth]{../talk-dms/figs/siaweaktime.png}}

@@

\subsection{the algorithm}

@@

\begin{itemize}
\item new algorithm:\footnote{E.~Bueler \& P.~Farrell (2024). \emph{A full approximation scheme multilevel method for nonlinear variational inequalities}, SIAM J.~Sci.~Comput.~46 (4) \sdoi{10.1137/23M1594200}}

\bigskip
{\color{FireBrick} FASCD = full approximation scheme constraint decomposition}

\bigskip
    \begin{itemize}
    \item[$\circ$] what is ``constraint decomposition''?
    \item[$\circ$] unilateral (lower obstacle) version shown, but bilateral or ``box constraints'' are implemented
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item start with nested meshes:
    $$\Omega^j \subset \Omega^{j+1}$$
\item the FE vector spaces $\mathcal{V}^j$ over $\Omega^j$ are also nested:
    $$\mathcal{V}^j \subset \mathcal{V}^{j+1}$$
\item $\mathcal{V}^J = \sum_{i=0}^J \mathcal{V}^i$ is called a \emph{subspace decomposition}

    \begin{itemize}
    \item[$\circ$] \emph{non}-unique vector space sum
    \item[$\circ$] one may analyze linear multigrid for PDEs via subspace decomposition
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Tai (2003): constraint decomposition \emph{non-trivially} extends subspace decomposition to convex subsets
\item suppose $\mathcal{K}^J \subset \mathcal{V}^J$ is a closed and convex subset
\item suppose $\mathcal{V}^J = \sum_{i=0}^J \mathcal{V}^i$ is a subspace decomposition
\begin{block}{definition}
$\ds \mathcal{K}^J = \sum_{i=0}^J \mathcal{K}^i$ \quad is a \emph{constraint decomposition} (CD) if there are closed and convex subsets $\mathcal{K}^i\subset \mathcal{V}^i$, and (nonlinear) projections $\Pi_i : \mathcal{K}^J \to \mathcal{K}^i$, so that $\ds v = \sum_{i=0}^J \Pi_i v$ and a stability condition applies (not shown)
\end{block}
\end{itemize}
\end{frame}

@@

\begin{frame}{iterations over a constraint decomposition}

\begin{itemize}
\item Tai proposed abstract iterations for solving $VI(F,\ell,\mathcal{K})$ over a CD
\item specifically the multiplicative/serial iteration:

@@

\item this iteration is not practical because you must evaluate all residuals on the finest level
\item two techniques make a practical solver:

    \begin{itemize}
    \item[$\circ$] \emph{defect obstacles} on each level  \hfill {\scriptsize Gr\"aser \& Kornhuber (2009)}
    \item[$\circ$] \emph{FAS coarse corrections}  \hfill {\scriptsize Bueler \& Farrell (2024)}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item all sets $\mathcal{U}^j$ and $\mathcal{D}^j$ include the origin
\end{itemize}
\end{frame}

@@

see Bueler \& Farrell (2024) for:
\begin{itemize}
\item generalization to upper and lower obstacles:
    $$\mathcal{K}^J = \{\underline{\psi}^J \le v^J \le \overline{\psi}^J\}$$
\item stopping criteria
    \begin{itemize}
    \item[$\circ$] evaluates CP/KKT conditions
    \end{itemize}
\item FMG cycle
\item details of $O(m_J)$ smoother: a few iterations of projected, reduced-space Newton with a few iterations of CG or GMRES, plus ILU preconditioning, on the linear step equations
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item FASCD = new multilevel solver for VI (free-boundary) problems
    \begin{itemize}
    \item[$\circ$] implemented in Python using the Firedrake library
    \item[$\circ$] observed optimality and actually fast on tested problems
    \end{itemize}
\item many things \textbf{{\color{FireBrick} to do}}:
    \begin{itemize}
    \item[$\circ$] implement in C within PETSc
    \item[$\circ$] prove convergence; compare (Reusken, 1988)
    \item[$\circ$] mesh adaptivity, better smoothers, parabolic examples, \dots
    \end{itemize}
\end{itemize}

\bigskip\bigskip
\begin{minipage}[T]{0.25\textwidth}
\vspace{0pt}

\includegraphics[width=0.8\textwidth]{figs/QRbuelerfarrell.png}

\scriptsize paper in

SIAM J.~Sci.~Comput.
\end{minipage}
\hfill
\begin{minipage}[T]{0.25\textwidth}
\vspace{0pt}

\hfill \includegraphics[width=0.8\textwidth]{figs/QRfascd.png}

\scriptsize \hfill code at bitbucket

\phantom{foo}
\end{minipage}
\end{frame}
